Time invariant analysis of epidemics with \pkg{EpiCompare}

### 2021-03-31 jss/EpiCompare-paper/EpiCompare-paper.Rmd
some section 2 edits plus add of plotly vis

## Motivating time-invariant analysis through \shannon{the reproduction number} $R_0$^[\ben{Is this still a relevant title for this section? Think about this header after the rewrite.}] {short-title="motivating through R0" #sec:r0}

Time-invariant analysis, as it appears in \pkg{EpiCompare}, \shannon{bypasses} many difficulties \shannon{in} comparing different epidemics. With time-invariant analysis, comparing the decades-long outbreak of HIV in the US to a 10 day outbreak of norovirus on a cruise ship is \sout{still} possible. Time-dependent problems can arise when estimating epidemiological parameters, including the reproduction number $R_0$. \sout{\ben{We will use $R_0$ to motivate the usefulness of time-invariant analysis in this section.}}^[\shannon{I don't think this is a necessary sentence}]

@@

$R_0$ is probably the most famous time-invariant numerical summary of an epidemic and is often associated with the Susceptible-Infectious-Recovered (SIR) model \citep{hethcote2000}.  $R_0$ is a one-number summary of a disease and is defined as the expected number of infections caused by a single infector who is added to a completely susceptible population \citep{anderson1992}.  \shannon{This definition concerns the \textit{generations} of new infections produced by $R_0$ but says nothing about the scale of time in which these new infections can occur.  However, data informing models are often temporally anchored (e.g. weekly case counts).  As a result, $R_0$ can be a difficult quantity to estimate and disentangle from time-based data.  For example, \cite{Gallagher2020} demonstrate how $R_0$ can be sensitive to time-based parameters such as the beginning and end of an epidemic, two quantities that generally do not have precise definitions.} \ben{[Ben says: \@Shannon could we expand on this difficulty / show it really is difficult with 1 more sentence? It might benefit from listing the difficulting in estimating a robust $R_0$ value (difficulties: time steps used, scale and "start" and final start, assumptions about population growth and underlying model fit).]}\shannon{[Shannon says I added the few sentences to more smoothly introduce r0]} To demonstrate the difficulty of discerning $R_0$ in a time-dependent analysis, we first introduce \citet{Kermack1927}’s SIR model. This model captures the transitions from one state to the next as a system of ordinary differential equations, where $N$ is the total number of individuals, $\beta$ is the rate of infection, and $\gamma$ is the rate of recovery,

@@

From this model, $R_0 = \beta/\gamma$, aka the ratio of the infection rate compared to the recovery rate. Since $\beta$ and $\gamma$ are both rates, the ratio of the two, $R_0$, is a time-invariant quantity.  Once $R_0$ is estimated, practioners can infer important epidemic quantities such as the total number of infections or the percent of a population needed to be vacccinated to stop the sustained spread of an epidemic.  Moreover, $R_0$ allows us to compare different diseases and different instances of outbreaks on the same scale.

\ben{[Ben says: It's unclear to me why we have a subtitle here - isn't it just more motivation of time-invariant anlysis with $R_0$? Also, I feel like the story is weak here. The point is to leverage $R_0$ to show the value of time-invariant analysis - this seems a bit more like just discussing properties of $R_0$. In the follow rewrite I use "[" and "]" to indicate that this is a section from your earlier draft.]}
\shannon{[Shannon says: Tried to tie this better to the previous part since it's no longer a new section.  also highlighted tie to TIA and r0.  I also wanted to bring the punch line (overlapping epidemics = same r0) closer to the beginning so those who don't want to slog through mathematical details can get the takeaway.]}

\shannon{Time-invariant analysis helps practicioners to more easily compare $R_0$ from different outbreaks.} For example, consider two epidemics generated from the Kermack and McKendrick SIR equations. Tne first epidemic has parameters $\beta_1, \gamma_1 = (0.8,0.4)$ and the second has $\beta_2, \gamma_2 = (0.64,0.32)$.  Both epidemics have populations of 1000 people with 10 individuals initially infected.  Additionally note that the two reproduction numbers are the same for each epidemic, $R_0 = 2=0.8/0.4 = 0.64/0.32$.  We plot the epidemics with traditional $state$ vs. $time$ plots. In Fig. \ref{fig:different-scales-standard} we show the time-based paths for the $S$, $I$, and $R$ states for the first 15 days of observed data.  In this time-variant view, we may believe that \ben{epidemic} 1 has a larger $R_0$ than \ben{epidemic} 2 because the peak of infection occurs more quickly than in Epidemic 2.  On the other hand, we may believe \ben{epidemic} 2 has a larger $R_0$ because its unclear if the  number of infections in that \sout{model}\ben{epidemic} has not yet peaked at time 15.  In this time-variant view, we cannot determine if one epidemic has larger value of $R_0$.

@@

```{r warning = FALSE, message = FALSE, echo = FALSE, fig.cap = "\\label{fig:different-scales-tern}Example of two epidemics with different $\\beta$ and $\\gamma$ paremeters but the same initial reproduction number $R_0$ = 2.  Both plots are generated from models with $N= 1000$ individuals with $S(0) = 990$ and $I(0) = 10$.  These are plotted in the time-invariant view, where we can see the number of susceptible, infectious, and recovered."}
ggplot(data = df %>%
         filter(t <= 15)) +
  geom_path(aes(x = S, y = I, z = R,
                     col = type)) +
  geom_point(aes(x = S, y = I, z = R,
                     col = type)) +
    coord_tern() + 
  theme_sir() +
  scale_color_manual(values = c("red", "blue"), name = "Epidemic") 
```

@@

$$
\psi = \left \{(S(t), I(t), R(t)): S, I, R \ge 0, S + I + R = N \right \}_{t\in[0,T]},
$$
where a mapping $\xi : s \to \mathbb{R}$ that is monotonically increasing would not change the definition of $\psi$, i.e. $\psi_\xi = \psi$ where :

@@

The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent result from \cite{Harko2014} and is shown in detail in Proof \ref{proof:thm}.  The consequence of Theorem \ref{thm:sir-scale} is that for two SIR models that have the same initial percent of individuals in each state and $R_0$ then for every point on the epidemic path of the first SIR model \ben{\sout{is also} can be mapped to} a point on the epidemic path of the second SIR model. \shannon{In other words, the two epidemics form the same filamental trajectory. For SIR models with similar initial state percentages, time-invariant analysis, time-invariant analysis allows practicioners to compare values of $R_0$ at a glance.}  \ben{[We've changed how we describe these two objects when they are related to the theorem - is that desirable? ]}\shannon{added that sentence to for that}

>>>>

## Motivating time-invariant analysis through the reproduction number $R_0$ {short-title="motivating through R0" #sec:r0}

Time-invariant analysis, as it appears in \pkg{EpiCompare}, \shannon{bypasses} many difficulties \shannon{in} comparing different epidemics. With time-invariant analysis, comparing the decades-long outbreak of HIV in the US to a 10 day outbreak of norovirus on a cruise ship is \sout{still} possible. Time-dependent problems can arise when estimating epidemiological parameters, including the reproduction number $R_0$. \sout{\ben{We will use $R_0$ to motivate the usefulness of time-invariant analysis in this section.}}^[\shannon{I don't think this is a necessary sentence.} \ben{I still think it adds value to the story and I'm not sure people really read section titles that are long.}]

@@

$R_0$ is probably the most famous time-invariant numerical summary of an epidemic and is often associated with the Susceptible-Infectious-Recovered (SIR) model \citep{hethcote2000}.  $R_0$ is a one-number summary of a disease and is defined as the expected number of infections caused by a single infector who is added to a completely susceptible population \citep{anderson1992}. \ben{This means that $R_0$ is a time-invariant parameter yet is estimated with time-based data, which can make it a [difficult quantity to estimate.]}  \ben{\sout{This definition concerns the \textit{generations} of new infections produced by $R_0$ but says nothing about the scale of time in which these new infections can occur.  However, data informing models are often temporally anchored (e.g. weekly case counts).}}\footnote{\ben{This was a bit unclear including defining a new term.   As a result, $R_0$ can be a difficult quantity to estimate and disentangle from time-based data.  For example,}}\shannon{ \cite{Gallagher2020} demonstrate\ben{s} how $R_0$ can be sensitive to time-based parameters such as the beginning and end of an epidemic, two quantities that generally}\ben{hard to define precisely.\sout{do not have precise definitions}}. To demonstrate the difficulty of discerning $R_0$ in \ben{\sout{a}other}\footnote{\ben{I change this so we don't confused readers that we're going show the impact in tools beyond just the estimation itself.}} time-dependent analysis, we first introduce \citet{Kermack1927}’s SIR model. This model captures the transitions from one state to the next as a system of ordinary differential equations, where $N$ is the total number of individuals, $\beta$ is the rate of infection, and $\gamma$ is the rate of recovery,

@@

From this model, $R_0 = \beta/\gamma$, aka the ratio of the infection rate compared to the recovery rate. \ben{From this definition, given 
\sout{Since}} $\beta$ and $\gamma$ are both rates, \ben{it should be clear that} the ratio of the two, $R_0$, is a time-invariant quantity.\footnote{\ben{I am trying to make it look like we are not repeating ourselves by saying $R_0$ is time-invariant.}}  Once $R_0$ is estimated, practitioners can infer important epidemic quantities such as the total number of infections or the percent of a population needed to be vacccinated to stop the sustained spread of an epidemic.  Moreover, $R_0$ allows us to compare different diseases and different instances of outbreaks on the same scale.

\ben{[Ben says: It's unclear to me why we have a subtitle here - isn't it just more motivation of time-invariant anlysis with $R_0$? Also, I feel like the story is weak here. The point is to leverage $R_0$ to show the value of time-invariant analysis - this seems a bit more like just discussing properties of $R_0$. In the follow rewrite I use "[" and "]" to indicate that this is a section from your earlier draft.]}
\shannon{[Shannon says: Tried to tie this better to the previous part since it's no longer a new section.  also highlighted tie to time-invariant analysis and $R_0$.  I also wanted to bring the punch line (overlapping epidemics = same r0) closer to the beginning so those who don't want to slog through mathematical details can get the takeaway.]}

\shannon{Time-invariant analysis helps practitioners to more easily compare $R_0$ from different outbreaks.} For example, consider two epidemics generated from the Kermack and McKendrick SIR equations. The first epidemic has parameters $\beta_1, \gamma_1 = (0.8,0.4)$ and the second has $\beta_2, \gamma_2 = (0.64,0.32)$.  Both epidemics have populations of 1000 people with 10 individuals initially infected.  Additionally note that the two reproduction numbers are the same for each epidemic, $R_0 = 2=0.8/0.4 = 0.64/0.32$.  We plot the epidemics with traditional $state$ vs. $time$ plots. In Fig. \ref{fig:different-scales-standard} we show the time-based paths for the $S$, $I$, and $R$ states for the first 15 days of observed data.  In this time-variant view, we may believe that \ben{epidemic} 1 has a larger $R_0$ than \ben{epidemic} 2 because the peak of infection occurs more quickly than in Epidemic 2.  On the other hand, we may believe \ben{epidemic} 2 has a larger $R_0$ because its unclear if the  number of infections in that \sout{model}\ben{epidemic} has not yet peaked at time 15.  In this time-variant view, we cannot determine if one epidemic has larger value of $R_0$.

@@

```{r warning = FALSE, message = FALSE, echo = FALSE, fig.cap = "\\label{fig:different-scales-tern}Example of two epidemics with different $\\beta$ and $\\gamma$ paremeters but the same initial reproduction number $R_0$ = 2.  Both plots are generated from models with $N= 1000$ individuals with $S(0) = 990$ and $I(0) = 10$.  These are plotted in the time-invariant view, where we can see the number of susceptible, infectious, and recovered.", fig.show="hold", out.width="49%"}
tern_vis <- ggplot(data = df %>%
         filter(t <= 15)) +
  geom_path(aes(x = S, y = I, z = R,
                     col = type)) +
  geom_point(aes(x = S, y = I, z = R,
                     col = type)) +
    coord_tern() + 
  theme_sir() +
  scale_color_manual(values = c("red", "blue"), name = "Epidemic")

df_tri <- data.frame(t = 0:3,
                     S = c(1,0,0,1),
                     I = c(0,1,0,0),
                     R = c(0,0,1,0),
                     name = c("S", "I", "R", "S"),
                     type = "border") %>% 
  mutate(type = as.character(type))
df2 <- df %>% filter(t <= 15) %>%
  mutate(N = S+I+R,
        S = S/N,
        I = I/N,
        R = R/N) %>%
  select(-N)

if (F){
  plotly_vis <- df2 %>% 
    plotly::plot_ly(x = ~S, y = ~R, z = ~I, color = ~type) %>%
    plotly::add_trace(type = 'scatter3d', mode = "lines+markers",
                      marker = list(size = 5),
                      line = list(width = 4)) %>%
    plotly::add_trace(data = df_tri, 
                      x = ~S, 
                      y = ~R,
                      z = ~I,
                      type = "mesh3d", color = I("blue"), opacity = .1) %>%
    plotly::add_text(data = df_tri[1:3,], 
                      x = ~S, 
                      y = ~R,
                      z = ~I, 
                     text= ~name, color = I("black")) %>%
    plotly::layout(scene = list(
        camera = list(eye = list(x = 1.5, y = 1.5, z = 1.2)),
        aspectratio = list(x = 1, y = 1, z = 1),
        xaxis = list(title = ""),
        yaxis = list(title = ""),
        zaxis = list(title = "")),
        showlegend = FALSE)
    #plotly::add_surface(data = df_tri, x = ~S, y = ~I, z = ~R, color = ~type)

  Sys.setenv("PATH" = paste(Sys.getenv("PATH"), 
                            "/Users/benjaminleroy/miniconda3/bin", 
                            sep = .Platform$path.sep))
  plotly::orca(plotly_vis, file = "images/vis3d.png",width = 8 * 120, height = 8 * 120)
}

knitr::include_graphics("images/vis3d.png")
plot(tern_vis)
```

@@

$$
\psi = \left \{(S(t), I(t), R(t)): S, I, R \ge 0, S + I + R = N \right \}_{t\in[0,T]},
$$
where a mapping $\xi : s \to \mathbb{R}$ that is strictly monotonically increasing would not change the definition of $\psi$, i.e. $\psi_\xi \equiv \psi$ where :

@@

The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent result from \cite{Harko2014} and is shown in detail in Proof \ref{proof:thm}.  The consequence of Theorem \ref{thm:sir-scale} is that for two SIR models that have the same initial percent of individuals in each state and $R_0$ then for every point on the epidemic path of the first SIR model \ben{\sout{is also} can be mapped to} a point on the epidemic path of the second SIR model. \shannon{In other words, the two epidemics form the same filamental trajectory. For SIR models with similar initial state percentages, time-invariant analysis allows practitioners to compare values of $R_0$ at a glance.}

### 2021-03-31 jss/EpiCompare-paper/EpiCompare-paper.tex
some section 2 edits plus add of plotly vis

\subsection[motivating through R0]{Motivating time-invariant analysis
through \textcolor{orange}{the reproduction number}
\(R_0\)\footnote{\textcolor{violet}{Is this still a relevant title for this section? Think about this header after the rewrite.}}}\label{sec:r0}

Time-invariant analysis, as it appears in \pkg{EpiCompare},
\textcolor{orange}{bypasses} many difficulties \textcolor{orange}{in}
comparing different epidemics. With time-invariant analysis, comparing
the decades-long outbreak of HIV in the US to a 10 day outbreak of
norovirus on a cruise ship is \sout{still} possible. Time-dependent
problems can arise when estimating epidemiological parameters, including
the reproduction number \(R_0\).
\sout{\textcolor{violet}{We will use $R_0$ to motivate the usefulness of time-invariant analysis in this section.}}\footnote{\textcolor{orange}{I don't think this is a necessary sentence}}

\(R_0\) is probably the most famous time-invariant numerical summary of
an epidemic and is often associated with the
Susceptible-Infectious-Recovered (SIR) model \citep{hethcote2000}.
\(R_0\) is a one-number summary of a disease and is defined as the
expected number of infections caused by a single infector who is added
to a completely susceptible population \citep{anderson1992}.
\textcolor{orange}{This definition concerns the \textit{generations} of new infections produced by $R_0$ but says nothing about the scale of time in which these new infections can occur.  However, data informing models are often temporally anchored (e.g. weekly case counts).  As a result, $R_0$ can be a difficult quantity to estimate and disentangle from time-based data.  For example, \cite{Gallagher2020} demonstrate how $R_0$ can be sensitive to time-based parameters such as the beginning and end of an epidemic, two quantities that generally do not have precise definitions.}
\textcolor{violet}{[Ben says: \@Shannon could we expand on this difficulty / show it really is difficult with 1 more sentence? It might benefit from listing the difficulting in estimating a robust $R_0$ value (difficulties: time steps used, scale and "start" and final start, assumptions about population growth and underlying model fit).]}\textcolor{orange}{[Shannon says I added the few sentences to more smoothly introduce r0]}
To demonstrate the difficulty of discerning \(R_0\) in a time-dependent
analysis, we first introduce \citet{Kermack1927}'s SIR model. This model
captures the transitions from one state to the next as a system of
ordinary differential equations, where \(N\) is the total number of
individuals, \(\beta\) is the rate of infection, and \(\gamma\) is the
rate of recovery,

@@

From this model, \(R_0 = \beta/\gamma\), aka the ratio of the infection
rate compared to the recovery rate. Since \(\beta\) and \(\gamma\) are
both rates, the ratio of the two, \(R_0\), is a time-invariant quantity.
Once \(R_0\) is estimated, practioners can infer important epidemic
quantities such as the total number of infections or the percent of a
population needed to be vacccinated to stop the sustained spread of an
epidemic. Moreover, \(R_0\) allows us to compare different diseases and
different instances of outbreaks on the same scale.

\textcolor{violet}{[Ben says: It's unclear to me why we have a subtitle here - isn't it just more motivation of time-invariant anlysis with $R_0$? Also, I feel like the story is weak here. The point is to leverage $R_0$ to show the value of time-invariant analysis - this seems a bit more like just discussing properties of $R_0$. In the follow rewrite I use "[" and "]" to indicate that this is a section from your earlier draft.]}
\textcolor{orange}{[Shannon says: Tried to tie this better to the previous part since it's no longer a new section.  also highlighted tie to TIA and r0.  I also wanted to bring the punch line (overlapping epidemics = same r0) closer to the beginning so those who don't want to slog through mathematical details can get the takeaway.]}

\textcolor{orange}{Time-invariant analysis helps practicioners to more easily compare $R_0$ from different outbreaks.}
For example, consider two epidemics generated from the Kermack and
McKendrick SIR equations. Tne first epidemic has parameters
\(\beta_1, \gamma_1 = (0.8,0.4)\) and the second has
\(\beta_2, \gamma_2 = (0.64,0.32)\). Both epidemics have populations of
1000 people with 10 individuals initially infected. Additionally note
that the two reproduction numbers are the same for each epidemic,
\(R_0 = 2=0.8/0.4 = 0.64/0.32\). We plot the epidemics with traditional
\(state\) vs.~\(time\) plots. In Fig.
\ref{fig:different-scales-standard} we show the time-based paths for the
\(S\), \(I\), and \(R\) states for the first 15 days of observed data.
In this time-variant view, we may believe that
\textcolor{violet}{epidemic} 1 has a larger \(R_0\) than
\textcolor{violet}{epidemic} 2 because the peak of infection occurs more
quickly than in Epidemic 2. On the other hand, we may believe
\textcolor{violet}{epidemic} 2 has a larger \(R_0\) because its unclear
if the number of infections in that
\sout{model}\textcolor{violet}{epidemic} has not yet peaked at time 15.
In this time-variant view, we cannot determine if one epidemic has
larger value of \(R_0\).

@@

{\centering \includegraphics{Figs/unnamed-chunk-3-1}

@@

\[
\psi = \left \{(S(t), I(t), R(t)): S, I, R \ge 0, S + I + R = N \right \}_{t\in[0,T]},
\] where a mapping \(\xi : s \to \mathbb{R}\) that is monotonically
increasing would not change the definition of \(\psi\),
i.e.~\(\psi_\xi = \psi\) where :

@@

The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent
result from \cite{Harko2014} and is shown in detail in Proof
\ref{proof:thm}. The consequence of Theorem \ref{thm:sir-scale} is that
for two SIR models that have the same initial percent of individuals in
each state and \(R_0\) then for every point on the epidemic path of the
first SIR model \textcolor{violet}{\sout{is also} can be mapped to} a
point on the epidemic path of the second SIR model.
\textcolor{orange}{In other words, the two epidemics form the same filamental trajectory. For SIR models with similar initial state percentages, time-invariant analysis, time-invariant analysis allows practicioners to compare values of $R_0$ at a glance.}
\textcolor{violet}{[We've changed how we describe these two objects when they are related to the theorem - is that desirable? ]}\textcolor{orange}{added that sentence to for that}

@@

Comparison and assessment of model fit or comparisons of one model to
another model can be performed in a variety of ways including mean
square error, AIC, plots, and more. Perhaps the most useful tool
\pkg{EpiCompare} has to offer to the expert, for comparison and
assessment of models, is in its post-processing tools which create a
standard output. It is then a matter of writing a script or function
made for that standard output to assess the results from multiple models
in the way the user
desires\footnote{\textcolor{violet}{[Ben says: Shannon, doesn't this below in the previous section?, I might drop this paragraph except for the introduction sentence - is it really true that the standardization is most valuable?]}}.

However, for those who like more concrete tools, \pkg{EpiCompare} offers
functions to compare prediction regions to one another including
\texttt{geom\_prediction\_band()} (which plots the region), and
\texttt{create\_\{convex\_hull,delta\_ball\}\_structure()}(which returns
the \texttt{R} output for the given structure), and \texttt{contained()}
(which allows the user to determine if one set of points is contained in
a prediction band). Additionally, we offer ways to determine if model
outputs are compatible with one another, that is how extreme one output
is to another. \textcolor{orange}{Ben says something about distance}

>>>>

\subsection[motivating through R0]{Motivating time-invariant analysis
through the reproduction number \(R_0\)}\label{sec:r0}

Time-invariant analysis, as it appears in \pkg{EpiCompare},
\textcolor{orange}{bypasses} many difficulties \textcolor{orange}{in}
comparing different epidemics. With time-invariant analysis, comparing
the decades-long outbreak of HIV in the US to a 10 day outbreak of
norovirus on a cruise ship is \sout{still} possible. Time-dependent
problems can arise when estimating epidemiological parameters, including
the reproduction number \(R_0\).
\sout{\textcolor{violet}{We will use $R_0$ to motivate the usefulness of time-invariant analysis in this section.}}\footnote{\textcolor{orange}{I don't think this is a necessary sentence.}
  \textcolor{violet}{I still think it adds value to the story and I'm not sure people really read section titles that are long.}}

\(R_0\) is probably the most famous time-invariant numerical summary of
an epidemic and is often associated with the
Susceptible-Infectious-Recovered (SIR) model \citep{hethcote2000}.
\(R_0\) is a one-number summary of a disease and is defined as the
expected number of infections caused by a single infector who is added
to a completely susceptible population \citep{anderson1992}.
\textcolor{violet}{This means that $R_0$ is a time-invariant parameter yet is estimated with time-based data, which can make it a [difficult quantity to estimate.]}
\textcolor{violet}{\sout{This definition concerns the \textit{generations} of new infections produced by $R_0$ but says nothing about the scale of time in which these new infections can occur.  However, data informing models are often temporally anchored (e.g. weekly case counts).}}\footnote{\textcolor{violet}{This was a bit unclear including defining a new term.   As a result, $R_0$ can be a difficult quantity to estimate and disentangle from time-based data.  For example,}}\textcolor{orange}{ \cite{Gallagher2020} demonstrate\textcolor{violet}{s} how $R_0$ can be sensitive to time-based parameters such as the beginning and end of an epidemic, two quantities that generally}\textcolor{violet}{hard to define precisely.\sout{do not have precise definitions}}.
To demonstrate the difficulty of discerning \(R_0\) in
\textcolor{violet}{\sout{a}other}\footnote{\textcolor{violet}{I change this so we don't confused readers that we're going show the impact in tools beyond just the estimation itself.}}
time-dependent analysis, we first introduce \citet{Kermack1927}'s SIR
model. This model captures the transitions from one state to the next as
a system of ordinary differential equations, where \(N\) is the total
number of individuals, \(\beta\) is the rate of infection, and
\(\gamma\) is the rate of recovery,

@@

From this model, \(R_0 = \beta/\gamma\), aka the ratio of the infection
rate compared to the recovery rate.
\textcolor{violet}{From this definition, given 
\sout{Since}} \(\beta\) and \(\gamma\) are both rates,
\textcolor{violet}{it should be clear that} the ratio of the two,
\(R_0\), is a time-invariant
quantity.\footnote{\textcolor{violet}{I am trying to make it look like we are not repeating ourselves by saying $R_0$ is time-invariant.}}
Once \(R_0\) is estimated, practitioners can infer important epidemic
quantities such as the total number of infections or the percent of a
population needed to be vacccinated to stop the sustained spread of an
epidemic. Moreover, \(R_0\) allows us to compare different diseases and
different instances of outbreaks on the same scale.

\textcolor{violet}{[Ben says: It's unclear to me why we have a subtitle here - isn't it just more motivation of time-invariant anlysis with $R_0$? Also, I feel like the story is weak here. The point is to leverage $R_0$ to show the value of time-invariant analysis - this seems a bit more like just discussing properties of $R_0$. In the follow rewrite I use "[" and "]" to indicate that this is a section from your earlier draft.]}
\textcolor{orange}{[Shannon says: Tried to tie this better to the previous part since it's no longer a new section.  also highlighted tie to time-invariant analysis and $R_0$.  I also wanted to bring the punch line (overlapping epidemics = same r0) closer to the beginning so those who don't want to slog through mathematical details can get the takeaway.]}

\textcolor{orange}{Time-invariant analysis helps practitioners to more easily compare $R_0$ from different outbreaks.}
For example, consider two epidemics generated from the Kermack and
McKendrick SIR equations. The first epidemic has parameters
\(\beta_1, \gamma_1 = (0.8,0.4)\) and the second has
\(\beta_2, \gamma_2 = (0.64,0.32)\). Both epidemics have populations of
1000 people with 10 individuals initially infected. Additionally note
that the two reproduction numbers are the same for each epidemic,
\(R_0 = 2=0.8/0.4 = 0.64/0.32\). We plot the epidemics with traditional
\(state\) vs.~\(time\) plots. In Fig.
\ref{fig:different-scales-standard} we show the time-based paths for the
\(S\), \(I\), and \(R\) states for the first 15 days of observed data.
In this time-variant view, we may believe that
\textcolor{violet}{epidemic} 1 has a larger \(R_0\) than
\textcolor{violet}{epidemic} 2 because the peak of infection occurs more
quickly than in Epidemic 2. On the other hand, we may believe
\textcolor{violet}{epidemic} 2 has a larger \(R_0\) because its unclear
if the number of infections in that
\sout{model}\textcolor{violet}{epidemic} has not yet peaked at time 15.
In this time-variant view, we cannot determine if one epidemic has
larger value of \(R_0\).

@@

{\centering \includegraphics[width=0.49\linewidth]{images/vis3d} \includegraphics[width=0.49\linewidth]{Figs/unnamed-chunk-3-2}

@@

\[
\psi = \left \{(S(t), I(t), R(t)): S, I, R \ge 0, S + I + R = N \right \}_{t\in[0,T]},
\] where a mapping \(\xi : s \to \mathbb{R}\) that is strictly
monotonically increasing would not change the definition of \(\psi\),
i.e.~\(\psi_\xi \equiv \psi\) where :

@@

The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent
result from \cite{Harko2014} and is shown in detail in Proof
\ref{proof:thm}. The consequence of Theorem \ref{thm:sir-scale} is that
for two SIR models that have the same initial percent of individuals in
each state and \(R_0\) then for every point on the epidemic path of the
first SIR model \textcolor{violet}{\sout{is also} can be mapped to} a
point on the epidemic path of the second SIR model.
\textcolor{orange}{In other words, the two epidemics form the same filamental trajectory. For SIR models with similar initial state percentages, time-invariant analysis allows practitioners to compare values of $R_0$ at a glance.}

@@

As introduced in Section \ref{sec:beyond-r0-sir} there's a lot of
potential for time-invariant tools to help compare and assess epidemics
and models/simulations. In \pkg{EpiCompare} we provide a set of
comparison and assessment tools for models that extend beyond the
standard performance metrics (e.g.~means squared error, AIC) and focus
on assessing the structural information the models capture. This
approaches work well on models where online one ``cycle'' of the
epidemic has occurred (no recovered individuals have been susceptible
again). Epidemics are complex objects, and we provide tools to create
prediction regions out of simulated epidemics. For 3-state epidemic
models, we provide the \texttt{ggplot}/\texttt{ggtern} extension
\texttt{geom\_prediction\_band()} which creates a prediction region
around the top \(1-\alpha\) proportion of the simulations (where the
simulations treated as filaments). In \pkg{Epicompare} we also provide
these prediction regions for epidemic models with with more than 3
states. With \texttt{create\_convex\_hull\_structure} and
`\texttt{create\_delta\_ball\_structure} that will create different
geometric representations of prediction regions and an assoicated
\texttt{contained} method that assesses if a new epidemic (seen as a
filament) is contained in the prediction region.

We also provide functions to calculate the extremeness of a true
epidemic compared to simulated epidemics through the equi-distance
filamental trajectory representation as mentioned in Section
\ref{sec:beyond-r0-sir}. Specifically, functions like
\texttt{distance\_psuedo\_density\_function} can calculate a
psuedo-density estimate of the true epidemic relative to simulated ones.
Functions \texttt{distance\_depth\_function} and
\texttt{local\_distance\_depth\_function} provide depth scores that
suggest how geometrically central an epidemic is to simulations.

\textcolor{violet}{[Ben says: Replaced with paragraphs above.]}
Comparison and assessment of model fit or comparisons of one model to
another model can be performed in a variety of ways including mean
square error, AIC, plots, and more. Perhaps the most useful tool
\pkg{EpiCompare} has to offer to the expert, for comparison and
assessment of models, is in its post-processing tools which create a
standard output. It is then a matter of writing a script or function
made for that standard output to assess the results from multiple models
in the way the user
desires\footnote{\textcolor{violet}{[Ben says: Shannon, doesn't this below in the previous section?, I might drop this paragraph except for the introduction sentence - is it really true that the standardization is most valuable?]}}.

\textcolor{violet}{[Ben says: Replaced with paragraphs above.]} However,
for those who like more concrete tools, \pkg{EpiCompare} offers
functions to compare prediction regions to one another including
\texttt{geom\_prediction\_band()} (which plots the region), and
\texttt{create\_\{convex\_hull,delta\_ball\}\_structure()}(which returns
the \texttt{R} output for the given structure), and \texttt{contained()}
(which allows the user to determine if one set of points is contained in
a prediction band). Additionally, we offer ways to determine if model
outputs are compatible with one another, that is how extreme one output
is to another. \textcolor{orange}{Ben says something about distance}
